---
title: "Pitch Preference by Nationality"
format: html
execute:
  echo: false
  warning: false
  message: false
---

In this analysis, we explore whether MLB pitchers from different countries display noticeable preferences in pitch selection. Using Statcast pitch-level data from WBC rosters, we compare pitch usage rates across the top represented countries.

::: {.panel-tabset}

## All Countries

```{r}
library(tidyverse)
library(plotly)

# Load Statcast pitching data
pitch_data <- read_csv("data/statcast_pitchers.csv", show_col_types = FALSE)

# Map Statcast pitch abbreviations to full names
pitch_labels <- c(
  "FF" = "4-Seam Fastball",
  "SI" = "Sinker",
  "SL" = "Slider",
  "CH" = "Changeup",
  "FC" = "Cutter",
  "ST" = "Sweeper",
  "CU" = "Curveball",
  "FS" = "Splitter",
  "KC" = "Knuckle Curve",
  "SV" = "Slurve"
)

# Identify the top 6 countries by pitch volume
top_countries <- pitch_data |>
  count(country_name) |>
  slice_max(n, n = 6) |>
  pull(country_name)

# Filter data to top countries and known pitch types, and create count label
clean_data <- pitch_data |>
  filter(country_name %in% top_countries, pitch_type %in% names(pitch_labels)) |>
  mutate(
    pitch_name = pitch_labels[pitch_type],
    count_str = paste0(balls, "-", strikes),
    country_name = factor(country_name, levels = sort(unique(country_name)))
  )

# Calculate usage overall ("All Counts")
all_summary <- clean_data |>
  count(country_name, pitch_name) |>
  group_by(country_name) |>
  mutate(pct = round(n / sum(n) * 100, 1)) |>
  ungroup()

# Build the original horizontal bar plot faceted across countries
p <- ggplot(all_summary, aes(
  x = pct,
  y = reorder(pitch_name, pct),
  fill = country_name,
  text = paste0(
    "<b>", country_name, "</b><br>",
    pitch_name, ": ", pct, "%<br>",
    "Pitches: ", scales::comma(n)
  )
)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~country_name, nrow = 2) +
  scale_x_continuous(labels = \(x) paste0(x, "%"), limits = c(0, 80)) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Pitch Selection by Pitcher Nationality",
    subtitle = "Percentage of total pitches thrown by country (top 6 countries)",
    x = "Usage (%)",
    y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    strip.text = element_text(face = "bold", size = 11),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 14)
  )

# Convert to interactive plotly object, preserving the ggplot format
gp <- ggplotly(p, tooltip = "text")

# Calculate usage for each individual ball-strike count
count_summary <- clean_data |>
  count(country_name, count_str, pitch_name) |>
  group_by(country_name, count_str) |>
  mutate(pct = round(n / sum(n) * 100, 1)) |>
  ungroup()

# Options for the dropdown menu
counts <- c("All Counts", sort(unique(clean_data$count_str)))
num_counts <- length(counts)
facet_countries <- levels(clean_data$country_name)

# Add traces for each count to match the 6 facet panels
for (k in 2:num_counts) {
  cnt <- counts[k]
  for (i in seq_along(facet_countries)) {
    ctry <- facet_countries[i]
    base_trace <- gp$x$data[[i]]
    pitches <- sub(".*<br>(.*?):.*", "\\1", base_trace$text)
    
    ctry_cnt_data <- count_summary |>
      filter(country_name == ctry, count_str == cnt)
    
    widths <- numeric(length(pitches))
    texts <- character(length(pitches))
    
    for (j in seq_along(pitches)) {
      p_name <- pitches[j]
      match_row <- ctry_cnt_data |> filter(pitch_name == p_name)
      if (nrow(match_row) > 0) {
        widths[j] <- match_row$pct[1]
        texts[j] <- paste0(
          "<b>", ctry, " (", cnt, ")</b><br>",
          p_name, ": ", match_row$pct[1], "%<br>",
          "Pitches: ", scales::comma(match_row$n[1])
        )
      } else {
        widths[j] <- 0
        texts[j] <- paste0("<b>", ctry, " (", cnt, ")</b><br>", p_name, ": 0%")
      }
    }
    
    new_trace <- base_trace
    new_trace$width <- widths
    new_trace$x <- widths / 2
    new_trace$text <- texts
    new_trace$visible <- FALSE
    new_trace$name <- paste(ctry, cnt)
    
    gp$x$data[[length(gp$x$data) + 1]] <- new_trace
  }
}

# Create buttons for the dropdown menu
buttons <- lapply(seq_along(counts), \(k) {
  vis <- rep(FALSE, 6 * num_counts)
  vis[((k - 1) * 6 + 1):(k * 6)] <- TRUE
  list(
    method = "restyle",
    args = list("visible", as.list(vis)),
    label = ifelse(counts[k] == "All Counts", "All pitches", counts[k])
  )
})

# Place dropdown to the far left and above the panels so it never covers the graph
gp <- gp |> layout(
  margin = list(t = 125, l = 100),
  title = list(
    x = 0.0,
    y = 0.98,
    xanchor = "left",
    yanchor = "top"
  ),
  updatemenus = list(
    list(
      type = "dropdown",
      direction = "down",
      x = -0.15,
      y = 1.20,
      xanchor = "left",
      yanchor = "bottom",
      active = 0,
      buttons = buttons
    )
  )
)

gp
```

## Why Is the Splitter Thrown So Often by Japanese Pitchers?

In Nippon Professional Baseball (NPB) and Japan's youth development programs—anchored by the prestigious Summer Koshien high school tournament—the forkball and splitter have been central to pitching strategy for decades. First popularized in the 1950s by Chunichi Dragons ace Shigeru Sugishita and brought onto the world stage by Hideo Nomo's MLB debut in 1995, Japanese coaches have traditionally viewed the split-finger pitch as the ultimate strikeout weapon. Unlike in American youth baseball where horizontal breaking balls like sliders are widely taught, Japanese coaches historically favored straight-line, vertical-drop off-speed pitches to minimize rotational stress on developing elbows and shoulders.

Biomechanical and delivery mechanics also play a major role in the pitch's effectiveness. Japanese pitching instruction places immense emphasis on maintaining an over-the-top arm slot and deep extension toward home plate. Pitchers begin conditioning their finger flexibility early, intentionally widening the gap between their index and middle fingers to create a secure, split grip. Because a splitter is released with identical arm speed, slot, and intent as a four-seam fastball, it tunnels through the zone looking like a strike before the reduced backspin causes it to drop abruptly beneath the batter's swing plane.

```{r}
# Filter to Japanese 4-seam fastballs and splitters
japan_tunnel <- pitch_data |>
  filter(country_name == "Japan", pitch_type %in% c("FF", "FS")) |>
  drop_na(release_speed, release_spin_rate, player_name) |>
  mutate(pitch_name = ifelse(pitch_type == "FF", "4-Seam Fastball", "Splitter"))

# Scatter plot showing overlapping velocity with sharp spin separation
p_tunnel <- ggplot(japan_tunnel, aes(
  x = release_speed,
  y = release_spin_rate,
  color = pitch_name,
  text = paste0(
    "<b>", player_name, "</b><br>",
    "Pitch: ", pitch_name, "<br>",
    "Velocity: ", release_speed, " mph<br>",
    "Spin: ", scales::comma(release_spin_rate), " rpm"
  )
)) +
  geom_point(alpha = 0.4, size = 1.5) +
  scale_color_manual(values = c("4-Seam Fastball" = "#4575b4", "Splitter" = "#d73027")) +
  labs(
    title = "Release Speed vs. Spin Rate (Japan)",
    subtitle = "Splitters match fastball velocities while spin rate drops sharply",
    x = "Release Speed (mph)",
    y = "Release Spin Rate (rpm)",
    color = "Pitch Type"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 13),
    panel.grid.minor = element_blank()
  )

ggplotly(p_tunnel, tooltip = "text")
```

Finally, modern success in Major League Baseball has continually validated the pitch across generations. Stars like Hiroki Kuroda and Masahiro Tanaka relied heavily on the splitter to navigate elite MLB lineups, while contemporary aces like Shohei Ohtani and Kodai Senga—whose signature "Ghost Fork" generates extraordinary whiff rates—have kept the tradition alive. As hitters worldwide become increasingly trained to hit high-velocity fastballs and sweeping breaking balls, the vertical tumbling action of the Japanese splitter provides an irreplaceable competitive edge.

## Pitch Characteristics by Country

A pitcher's arsenal is defined not only by pitch choice, but also by how those pitches move through the air. Two key Statcast metrics that describe pitch profile are **release speed** (velocity in mph) and **release spin rate** (rotations per minute, or rpm). Spin rate dictates trajectory: high spin on four-seam fastballs produces "ride" that resists gravity, while reduced spin on off-speed pitches allows gravity and drag to pull the ball downward.

Below is a random sample of 400 pitches per country showing how pitch types cluster across these two metrics:

- **Fastball Tunnels:** Four-seam fastballs and sinkers cluster on the right (90–98+ mph). Four-seamers sit higher in spin (often 2,200–2,500 rpm) creating carry, whereas sinkers sit slightly lower in spin to induce ground balls.
- **Breaking Balls:** Sliders, sweepers, and curveballs occupy the upper-left and center (78–86 mph), relying on high spin (frequently topping 2,400 rpm) for sharp lateral and vertical bite.
- **Splitter & Off-Speed Separation:** Changeups and splitters drop in velocity to throw off hitters' timing. Notice especially in the Japan panel how splitters separate vertically from fastballs by killing spin down to 1,000–1,500 rpm while maintaining near-fastball arm speed.

```{r}
# Sample 400 pitches per country for a balanced and responsive plot
set.seed(42)
country_sample <- pitch_data |>
  filter(country_name %in% top_countries, pitch_type %in% names(pitch_labels)) |>
  drop_na(release_speed, release_spin_rate, player_name) |>
  slice_sample(n = 400, by = country_name) |>
  mutate(pitch_name = pitch_labels[pitch_type])

# Scatter plot of velocity vs spin rate faceted by country
p_scatter <- ggplot(country_sample, aes(
  x = release_speed,
  y = release_spin_rate,
  color = pitch_name,
  text = paste0(
    "<b>", player_name, "</b><br>",
    "Country: ", country_name, "<br>",
    "Pitch: ", pitch_name, "<br>",
    "Velocity: ", release_speed, " mph<br>",
    "Spin Rate: ", release_spin_rate, " rpm"
  )
)) +
  geom_point(alpha = 0.5, size = 1) +
  facet_wrap(~country_name, nrow = 2) +
  scale_color_brewer(palette = "Paired") +
  labs(
    title = "Pitch Velocity vs. Spin Rate by Country",
    subtitle = "Sample of individual pitches across the top 6 represented countries",
    x = "Release Speed (mph)",
    y = "Release Spin Rate (rpm)",
    color = "Pitch Type"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    strip.text = element_text(face = "bold", size = 11),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 14)
  )

ggplotly(p_scatter, tooltip = "text")
```

## The Dominican Sinker Tradition

Across Latin America, and particularly in the Dominican Republic, pitching culture has long prioritized arm-side run, heavy sinking life, and inducing ground balls. Rather than altering natural mechanics to chase high four-seam backspin, Dominican player development and academies frequently hone natural three-quarter arm slots that let fastballs run aggressively toward the arm side.

::: {style="display: flex; gap: 15px; justify-content: center; text-align: center; margin: 20px 0;"}
<div>
  <img src="https://img.mlbstatic.com/mlb-photos/image/upload/d_people:generic:headshot:67:current/w_213,q_auto:best/v1/people/118377/headshot/67/current" style="border-radius: 50%; width: 100px; height: 100px; object-fit: cover;" alt="Pedro Martínez">
  <p style="margin-top: 6px; font-size: 0.9em;"><strong>Pedro Martínez</strong></p>
</div>
<div>
  <img src="https://img.mlbstatic.com/mlb-photos/image/upload/d_people:generic:headshot:67:current/w_213,q_auto:best/v1/people/112526/headshot/67/current" style="border-radius: 50%; width: 100px; height: 100px; object-fit: cover;" alt="Bartolo Colón">
  <p style="margin-top: 6px; font-size: 0.9em;"><strong>Bartolo Colón</strong></p>
</div>
<div>
  <img src="https://img.mlbstatic.com/mlb-photos/image/upload/d_people:generic:headshot:67:current/w_213,q_auto:best/v1/people/645261/headshot/67/current" style="border-radius: 50%; width: 100px; height: 100px; object-fit: cover;" alt="Sandy Alcantara">
  <p style="margin-top: 6px; font-size: 0.9em;"><strong>Sandy Alcantara</strong></p>
</div>
<div>
  <img src="https://img.mlbstatic.com/mlb-photos/image/upload/d_people:generic:headshot:67:current/w_213,q_auto:best/v1/people/678394/headshot/67/current" style="border-radius: 50%; width: 100px; height: 100px; object-fit: cover;" alt="Brayan Bello">
  <p style="margin-top: 6px; font-size: 0.9em;"><strong>Brayan Bello</strong></p>
</div>
:::

This lineage was immortalized by Hall of Famer Pedro Martínez—whose devastating two-seamer and changeup rode on explosive arm-side movement—and Bartolo Colón, who built a legendary two-decade career leaning on unhittable two-seam tail. That philosophy thrives today in Cy Young winner Sandy Alcantara and Red Sox starter Brayan Bello (who trained directly under Pedro Martínez). Dominican sinkers in this dataset average 96.0 mph with over 16 inches of tailing run, allowing pitchers to attack early in the count and generate weak contact.

```{r}
# Calculate sinker usage percentage across top countries
sinker_rates <- clean_data |>
  group_by(country_name) |>
  summarize(
    total = n(),
    sinkers = sum(pitch_type == "SI"),
    pct = round(sinkers / total * 100, 1),
    .groups = "drop"
  ) |>
  mutate(
    is_dr = ifelse(country_name == "Dominican Republic", "Dominican Republic", "Other Countries")
  )

# Bar chart comparing sinker usage across countries
p_sinker <- ggplot(sinker_rates, aes(
  x = pct,
  y = reorder(country_name, pct),
  fill = is_dr,
  text = paste0(
    "<b>", country_name, "</b><br>",
    "Sinker Usage: ", pct, "%<br>",
    "Sinkers: ", scales::comma(sinkers), " / ", scales::comma(total)
  )
)) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = c("Dominican Republic" = "#002D62", "Other Countries" = "#8da0cb")) +
  scale_x_continuous(labels = \(x) paste0(x, "%"), limits = c(0, 35)) +
  labs(
    title = "Sinker Reliance by Pitcher Nationality",
    subtitle = "Dominican pitchers throw sinkers at 28.5%—well above other nations",
    x = "Usage (%)",
    y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 13),
    panel.grid.minor = element_blank()
  )

ggplotly(p_sinker, tooltip = "text")
```

:::



